---
title: "Using R for Data Science: In-Course Assessment"
author: "Giulia Guasoni"
date: "2024-12-02"
output: word_document
font: Arial
fontsize: 11pt
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction
  Forty genes, derived from a set of 200, were analysed via descriptive statistics. This report will describe key aspects of the code used in data handling and highlight any emerging trends. 

```{r, include=FALSE}
#Installing and loading any necessary packages

#Make sure the relevant files are in the working directory
path <- setwd(getwd())

library(ggplot2)
library(tidyverse)
library(pheatmap)
library(tools)
```


```{r Loading_the_data, include = FALSE}
#The code was adapted from the following site(s):
#https://stackoverflow.com/questions/17499013/how-do-i-make-a-list-of-data-frames
```

# Methods: Defensive Coding
## Loading the Data Files
  A function was created to ensure that all the relevant files were converted to data frames that could later be manipulated in RStudio. If no .csv or .txt files were present in the directory, the function would stop and return a message. 
```{r Loading_the_data_continued, include = TRUE}
#Loading the data sets and creating variables
#Creating a list vector of all the .csv and .txt files
relevant_data  <- list.files(pattern = "\\.(csv|txt)$")

#Creating a function that will read the files in 'relevant_data' and turn them into data frames
create_dataframes <- function(relevant_data){
  if (length(relevant_data) == 0){
    stop("No data to analyse.")
  } else if (length(relevant_data) > 0) {
    for (i in seq_along(relevant_data)){
      if (grepl("\\.csv", relevant_data[i])) {
        message(paste0(relevant_data[i], " is a .csv") )
        assign(paste0(file_path_sans_ext(relevant_data[i]), "_df"), read.csv(relevant_data[i]), envir = .GlobalEnv)
      } else if (grepl("\\.txt", relevant_data[i])){
        message(paste0(relevant_data[i], "is a .txt") )
        assign(paste0(file_path_sans_ext(relevant_data[i]), "_df"), read.delim(relevant_data[i]), envir = .GlobalEnv)
      }}}
dataframes_list <- ls(pattern = "_df", envir = .GlobalEnv)
dataframes_list <- assign("dataframes_list", dataframes_list, envir = .GlobalEnv)
message("All the relevant data has been loaded. The data frames for this project are: ", paste(dataframes_list, collapse = ", "), ".")
  }

#Running the function
create_dataframes(relevant_data) 
```

##Preparing for Data Manipulation
  To ensure that the recently converted data frames are ready for data manipulation, two 'for' loops check for the presence of missing values and duplicate rows. If any are found, the data frames are cleaned and an appropriate message is sent to the console.
```{r preparation_for_manipulation, include = TRUE}
#Ensuring the data frames contain no empty cells
for (dataframe in dataframes_list){
  if (sum(is.na(dataframe)) > 0){
    message(sum(is.na(dataframe)), " missing value(s) have been found and the relevant columns/rows have been removed. The data frame is now ready for manipulation. Proceed if you are satisfied.")
    #Remove columns containing missing values
    dataframe <- dataframe[ , colSums(is.na(dataframe))==0]
  } else if (sum(is.na(dataframe)) == 0){
    message(dataframe, " is ready for manipulation.")
  }
}

#Ensuring the dataframes don't contain duplicates
for (name_df in dataframes_list){
  df = get(name_df)
  duplicates = sum(duplicated(df))
  if (duplicates > 0){
    message(sum(duplicated(df)), " duplicate(s) have been found in ", name_df, " and the relevant rows have been removed.", "\n", "The data frame is now ready for manipulation. Proceed if you are satisfied.")
    df <- unique(df)
    assign(name_df, df, envir = .GlobalEnv)
  } else if (duplicates == 0){
    message(name_df, " is ready for manipulation.")
  }
}
```


```{r data_manipulation, include=FALSE}
#Creating a new data frame with all the information

#The following code was adapted from the following websites: 
#https://dplyr.tidyverse.org/reference/across.html
#https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/numeric

#Binding together the data_all and gene_annotations data frames
total <- merge(gene_annotation_df, data_all_df, by="X") 
#Sorting the assigned genes in numerical order for ease of reading and manipulation
gene_list_sorted <- genelist_9_df[order(genelist_9_df$x), ]
#Matching the gene names in gene_list to the names in the 'total' data frame and then creating a new data frame with all the relevant information
matched_genes <- match(gene_list_sorted, total$X)
sample <- total[matched_genes, ]
#Remove the X column from the 'sample' data frame since it contains the same values as the Gene column
sample = subset(sample, select = -c(X))
#Set the values in the Gene column as characters (relevant for next step)
sample$Gene = as.character(sample$Gene)
#log-transform the values in the treatment columns
sample_logged <- mutate(sample, across(where(is.numeric), log10))
```

```{r data_formatting, include=FALSE}
#Changing the data formatting
#The following code was adapted from the following website:
#https://tidyr.tidyverse.org/reference/pivot_longer.html

#create a vector containing the column names for the treatments
treatment <- c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J", "K", "L")
#converting the data type
sample_logged_long <- pivot_longer(sample_logged, cols = all_of(treatment), names_to = "treatment", values_to = "expression")
#Adding the TreatmentGroup column to the new data frame. We're repeating the column 40 times because we have 480 rows in the sample_logged_long data frame.
treatment_group <- rep(sample_annotation_df$TreatmentGroup, times = 40)
sample_logged_long$TreatmentGroup <- as.character(treatment_group)
```

## Visualising Gene Expression via Scatterplot
  After data handling, a scatterplot (Fig. 1) was created to illustrate potential trends in gene expression. The plot indicates that Type XA (green) and XC (pink) genes have similar expression levels, whereas XB (blue) genes generally exhibit lower expression levels. Notable outliers of the data set are 1422659_at and 1439006_x_at, both Type XB genes, which present notably lower gene expression levels.

```{r scatterplot, fig.cap = "Figure 1:A scatter plot demonstrating log(expression) for each gene", fig.width= 8, fig.align='center', echo=FALSE}
#Plotting the data
#The code was adapted from the following websites
#https://r-charts.com/correlation/scatter-plot-group-ggplot2/
sample_logged_long$comb <- paste0(sample_logged_long$Type, sample_logged_long$LongName)

ggplot(sample_logged_long, aes(x = comb, y = expression, colour = Type, shape = TreatmentGroup)) +
  geom_point() +
  scale_color_manual(values = c("XA" = "#39FF39", 
                                "XB" = "#66CCFF", 
                                "XC" = "#FF79C6")) +
  theme(legend.position = "right") +
          theme_bw() +
          theme(panel.grid.major = element_blank()) +
         theme(panel.grid.minor = element_blank()) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
          labs(
            title = "log(expression) (vertical) for each gene (horizontal)."
            ) +
          xlab("Treatment Group") +
          ylab("log(Expression)") +
          theme(plot.title = element_text(size=12))
```

## Clustering Data Based on Genes and Treatment
  While the scatterplot helps compare trends in gene expression between gene types, it is less effective at revealing any differences between treatments. The heatmap below clusters the treatments and genes based on gene expression (Fig. 2). Treatments 1, 2, and 3 have similar effects on gene expression, whereas Treatment 4 stands isolated, showing distinct opposite effects. Further analysis of gene clusters revealed that Treatments 1, 2, and 3 typically increase the expression of Type XA and XB genes, whereas Treatment 4 predominantly increases the expression of Type XC genes. Interestingly, the first cluster of genes (1439006_x_at, 144144_at, and 1422659_at) all display a reduced expression under all four treatment types, though this effect is less obvious under Treatment 4. Two of these three genes are the same outliers observed in the scatterplot (Fig. 1). Overall, this heatmap shows that Treatments 1, 2, and 3 have similar effects on Type XA and Type XB genes, whereas Treatment 4 has seemingly converse effects.
	
```{r heatmap1, fig.cap = "Figure 2: Heatmap where data was clustered based on genes and treatments", fig.height= 6, fig.align = 'center', echo = FALSE}
#Create two heatmaps:
#a) one where both the genes and the samples are clustered 

#Creating a matrix file only containing the log(expression) values and the LongName column as row names
no_Gene <- subset(sample_logged, select = -c(Gene, Type))
rownames(no_Gene) <- no_Gene$LongName
no_Gene <- subset(no_Gene, select = -c(LongName))
test <- as.matrix(no_Gene)

#Creating metadata data frames to annotate the heatmaps with Treatment Type and Gene Type.
metadata_c <- as.data.frame(c(sample_logged_long$TreatmentGroup), row.names = sample_logged_long$treatment)
names(metadata_c)[1] <- 'Treatment'
metadata_r <- as.data.frame(c(sample_logged_long$Type), row.names = sample_logged_long$LongName)
names(metadata_r)[1] <- 'Gene Type'

#Colour-coding the Treatment and Gene Type values
annotationscolours <- list( "Treatment" = c("1" = "#B572C7",
                                            "2" = "#9B4D96", 
                                            "3" = "#6A1E9C",
                                            "4" = "#5A137C"),
                            "Gene Type" = c("XA" = "#39FF39", 
                                            "XB" = "#66CCFF", 
                                            "XC" = "#FF79C6"))
#Creating the heatmap
pheatmap(test, 
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         main = "Heatmap Where Both Genes and Treatment are Clustered",
         annotation_legend = TRUE,
         annotation_col = metadata_c,
         annotation_row = metadata_r,
         fontsize = 7,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         annotation_colors = annotationscolours)
```

## Clustering Data Only Based on Genes
  A second heatmap (Fig. 3) where only genes were clustered was created to expand on some of the potential relationships observed in the first heatmap. This figure reveals that the expression of genes 1439006_x_at, 144144_at, and 1422659_at increases from Treatments 1 to 4. In the subsequent clusters of genes, expression increases across Treatments 1 to 3, before plumetting under Treatment 4. A converse trend is observed in the final two clusters of genes, where expression gradually increases from Treatments 1 to 3, before spiking under Treatment 4. 
  The trends revealed by this heatmap suggest that the four treatments are serial dilutions. However, in the absence of contextual information regarding the functions of these treatments, it is difficult to determine the direction of concentration. If these treatments are meant to suppress the expression of genes such as 1439006_x_at, 144144_at, and 1422659_at, it is likely that concentration decreases from Treatments 1 to 4. However, if these treatments are meant to enhance the expression of the other 37 genes, the heatmap suggests that the concentration increases from Treatments 1 to 4. However, the figure illustrates an alternative expression pattern under Treatment 4 that conflicts with this idea. If there is a concentration threshold, beyond which alternative pathways are activated, it is possible that Treatment 4 is the most concentrated of the four treatments, as demonstrated by its unique effects on gene expression. 
	Ultimately, it is difficult to make conclusions regarding the exact effects of these treatments on gene expression without additional information regarding the treatments. 


```{r heatmap2, fig.cap = "Figure 3: A heatmap where data was only clustered based on genes", fig.height = 6, fig.align = 'centre', echo = FALSE}
#b) one where only the genes are clustered.
pheatmap(test,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         main = "Heatmap Where Only Genes are Clustered",
         annotation_legend = TRUE,
         annotation_col = metadata_c,
         annotation_row = metadata_r,
         fontsize = 7,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         annotation_colors = annotationscolours)
```

## Representing Distributions via Boxplot

  The boxplot (Fig. 4) below further illustrates the distribution of the data. First glimpses of the boxplot show that there is greater variation in the expression of Type XB genes (blue) than the other two gene types, and the only extreme outliers belong to Type XB, which is consistent with the findings seen in the scatterplot and heatmaps. In addition, genes expression of Type XC genes (pink) is higher under Treatment 4, which is also consistent with the findings from the heatmaps. An interesting finding from the boxplot is that Treatment 4 is the only treatment without any extreme outliers, and that the median expression of genes under Treatment 4 is slightly higher than under the other three treatments.


```{r boxplot, fig.cap = "Figure 4: A boxplot demonstrating the log(expression) of genes in response to Treatments 1, 2, 3, and 4. Jitterplots of gene types XA (green), XB (blue), and XC (pink) are overlayed on top of the boxplots",  echo=FALSE}
#Create a boxplot with an overlaying jitter-plot to represent the data
colours <- c("1" = "#B572C7", "2" = "#9B4D96", "3" = "#6A1E9C", "4" = "#5A137C")

boxplot(sample_logged_long$expression ~ sample_logged_long$TreatmentGroup, data = sample_logged_long, col = colours, main = "Figure 4: log(expression) of genes in response to Treatments 1, 2, 3, and 4. Jitterplots of gene types XA (green), XB (blue), and XC (pink) are overlayed on top of the boxplots", par(cex.main = 0.85), xlab = "Treatment Group", ylab ="log(Expression)")

substrip <- sample_logged_long[sample_logged_long$Type == "XA",]
stripchart(substrip$expression ~ substrip$TreatmentGroup, 
           data = substrip,
           method = "jitter",
           pch = 20,
           col = "#39FF39",
           vertical = TRUE,
           add = TRUE,
           at = c(0.7, 1.7, 2.7, 3.7))

substrip <- sample_logged_long[sample_logged_long$Type == "XB",]
stripchart(substrip$expression ~ substrip$TreatmentGroup, 
           data = substrip,
           method = "jitter",
           pch = 20,
           col = "#66CCFF",
           vertical = TRUE,
           add = TRUE,
           at = c(1, 2, 3, 4))

substrip <- sample_logged_long[sample_logged_long$Type == "XC",]
stripchart(substrip$expression ~ substrip$TreatmentGroup, 
           data = substrip,
           method = "jitter",
           pch = 20,
           col = "#FF79C6",
           vertical = TRUE,
           add = TRUE,
           at = c(1.3, 2.3, 3.3, 4.3))
```

## Limitations
  In the absence of contextual information regarding the gene types or treatment functions, it is difficult to make concrete conclusions regarding the effects of the various treatments on gene expression. In addition, comparisons are made more difficult due to the absence of a control group where no treatment is applied to the genes. 